---
title: "Basic Prediction of Football Stats"
author: "Karl Wenzel"
date: "August 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we shall try to predict how many yards a running back will run over the course of an entire season.  We shall use the players' previous season total yards amount as the primary independent variable, and the current year total yards as the dependent variable.  We shall also consider the number of games played as independent variables for our model as we progress.

```{r RB-Prediction}

library(plyr)

PRB = data.frame(read.csv("C:/Users/karl/Documents/Fantasy-Football/data/tidy/full/PRB.csv"))

a = PRB[ PRB$wk <= 17, ]
a = ddply(a, .(player,seas), summarize, yds = sum(yds, na.rm=TRUE), tds = sum(td, na.rm=TRUE), games = length(unique(gid)) )
a$avg.yds = a$yds / a$games

# 'a' shows yearly yards and td amounts per player per season.  We also have an average yards per game.
print(head(a))

b = data.frame( player=factor() )

for (s in 2001:2015) {
  
  a1 = a[ a$seas == (s-1) & a$yds > 500, c("player", "yds", "tds", "games", "avg.yds") ]
  a2 = a[ a$seas == s, c("player", "yds", "tds", "games", "avg.yds") ]  
  
  names(a1)[2] = "prev.yds"
  names(a1)[3] = "prev.tds"
  names(a1)[4] = "prev.games"
  names(a1)[5] = "prev.avg.yds"  
  names(a2)[2] = "curr.yds"
  names(a2)[3] = "curr.tds"
  names(a2)[4] = "curr.games"
  names(a2)[5] = "curr.avg.yds"  
  
  bx = join( a1, a2, by="player", type="inner")
  bx$seas = s  
  b = rbind( b, bx)
}

# 'b' gives us a juxtaposition of the previous yards/tds and currnet yards/tds for each player by season
print(head(b))
```

Now that we have our data ready for analysis, we shall apply some linear models.  Also, here are a few refresher links for your convenience:
- [Understanding lm() Output from Montana State University](http://www.montana.edu/screel/Webpages/conservation%20biology/Interpreting%20Regression%20Coefficients.html#/4)
- [Linear Models from Princeton University](http://data.princeton.edu/R/linearModels.html)

```{r RB-Prediction-LM}
# 1st attempt: linear model simply uses previous yards total to predict current yards total

c1 = lm( formula = curr.yds ~ prev.yds, data = b )
c1.percent = round(summary(c1)$r.squared * 100, digits=2)
print(paste(c1.percent, "% of variance explained by model.", sep=""))
      
# 2nd attempt: predict average yards per game played instead of total yards (hopefully reduce impact of injuries)

c2 = lm( formula = curr.avg.yds ~ prev.avg.yds, data = b )
c2.percent = round(summary(c2)$r.squared * 100, digits=2)
print(paste(c2.percent, "% of variance explained by model.", sep=""))
      
# 3rd attempt: given know prev total yards, prev games played, number of current games played, predict current yards

c3 = lm( formula = curr.yds ~ prev.yds + prev.games + curr.games, data = b)
c3.percent = round(summary(c3)$r.squared * 100, digits=2)
print(paste(c3.percent, "% of variance explained by model.", sep=""))

# The final attempt does a pretty decent job trying to predict, however it shows how
# important it is to know properly guess many games a player will play in the current season.
# Here's more details of the linear model:

print(summary(c3))
plot(c3)

```


